---
title: "Final Project Code and Figures"
author: "Katrina Liu"
date: '2022-12-10'
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(MASS)
```

# BMI 715 Final Project


```{r}
# Loading the NHANES Dataset
nhanes = read.csv("nhanes_13_14_subset_updated.csv", row.names = "X")
```


## Data Processing and Exploratory Analysis

### Remove Invalid Data Point


```{r}
# Filter the cohort with hypertension
nhanes_ht = nhanes %>% filter(BPQ020 == 2, BPXDI1>0, BPXSY1>0, RIDAGEYR<80) 
nhanes_ht_related = nhanes_ht %>% 
  dplyr::select(c("BMXBMI", "RIDAGEYR", "LBXTC","LBXSGL","LBXSTP",
                  "BPXSY1","BPXML1","BPXPLS", "BPXDI1", "BPQ100D",
                                                  "RIAGENDR")) %>% 
  drop_na %>% mutate(BPQ100D = BPQ100D-1, RIAGENDR = RIAGENDR-1)
head(nhanes_ht_related)
```

### Scaling the Predictors


```{r}
# Scale continuous variables
nhanes_ht_ready = nhanes_ht_related[1:9] %>% scale %>% as.data.frame
nhanes_ht_ready = cbind(nhanes_ht_ready, nhanes_ht_related[10], 
                        nhanes_ht_related[11])

```

### Relationships Between Variables


```{r, fig.height = 8, fig.width = 10}
# Plot collinearity and correlation between predictors
plot(nhanes_ht_ready[,-6], 
     main="Relationships between Each Pair of Predictors")
heatmap(cor(nhanes_ht_ready[,-6]),
        main = "Correlation coefficients of Each Pair of Predictors",Colv=NA, 
        Rowv=NA)

```

## Linear Regression Model

### Assessing the Relationship between the Predictors and the Outcome

```{r, fig.height = 8, fig.width = 10}
# Plot collinearity and coefficient of all variables
plot(nhanes_ht_ready, main="Relationships between Each Pair of Variables")
```

```{r, fig.height = 8, fig.width = 10}
# Assees the linear relationship between BPXSY1 and BPXML1
plot(lm(BPXSY1~BPXML1, data=nhanes_ht_ready))
```


```{r, fig.height = 8, fig.width = 10}
# Full model
lm.full = lm(BPXSY1~.,data=nhanes_ht_ready)
summary(lm.full)
```


```{r, fig.height = 8, fig.width = 10}
# Assess the full model
plot(lm.full)
```


### Variable Selection


```{r}
# Base model
lm.base = lm(BPXSY1~1,data=nhanes_ht_ready)
summary(lm.base)
```

```{r}
# Step-wise model selection
lm.step =stepAIC(lm.full, scope=list(lm.base, lm.full), direction = "both")
summary(lm.step)
```


## Evaluate Model Fit

### Plotting the Residuals

```{r, fig.height = 8, fig.width = 10}
# Residual plot
plot(fitted(lm.step), resid(lm.step), main="Residual Plot")
```



```{r}
library(Metrics)
library(caret)

# Partition training/testing data
random_sample = createDataPartition(nhanes_ht_ready$BPXSY1, p=0.8, list=FALSE)
training_data = nhanes_ht_ready[seq(random_sample),]
testing_data = nhanes_ht_ready[-seq(random_sample),]

# Train selected model and predict
lm.train = lm(BPXSY1~RIDAGEYR+RIAGENDR+BPXDI1+BPXML1, data=training_data)
predictions = predict(lm.train, testing_data)

# Output results
data.frame( RMSE = rmse(predictions, testing_data$BPXSY1),
            MAE = mae(predictions, testing_data$BPXSY1))
```


```{r}
# Train full model and predict
lm.train.full = lm(BPXSY1~., data=training_data)
full_predictions = predict(lm.train.full, testing_data)

data.frame( RMSE = rmse(full_predictions, testing_data$BPXSY1),
            MAE = mae(full_predictions, testing_data$BPXSY1))
```


## Alternative Models: Regularization


```{r, fig.height = 8, fig.width = 10}
library(glmnet)
# Process the ht as a matrix
ht_mat = nhanes_ht_ready %>% as.matrix

# Perform different penalties
cv_model_lasso = cv.glmnet(ht_mat[,-6], ht_mat[,6], nfolds=10,alpha=1)
cv_model_ridge = cv.glmnet(ht_mat[,-6], ht_mat[,6], nfolds=10,alpha=0)
cv_model_elastic = cv.glmnet(ht_mat[,-6], ht_mat[,6], nfolds=10,alpha=0.5)

# Plot the results
plot(cv_model_lasso, 
     main="Performance of the Regularized Regression with L1 Penalty")
plot(cv_model_ridge,
     main="Performance of the Regularized Regression with L2 Penalty")
plot(cv_model_elastic, 
     main="Performance of the Regularized Regression with Elastic Net")
```


```{r}
train_mat = training_data %>% as.matrix
test_mat = testing_data %>% as.matrix

# Train regularized regression model with optimal lambda
cv_final_lasso = glmnet(train_mat[,-6], train_mat[,6], nfolds=10,
                        alpha=1, lambda=cv_model_lasso$lambda.min)
cv_final_ridge = glmnet(train_mat[,-6], train_mat[,6], nfolds=10,
                        alpha=0, lambda=cv_model_ridge$lambda.min)
cv_final_elastic = glmnet(train_mat[,-6], train_mat[,6], nfolds=10,
                          alpha=0.5, lambda=cv_model_elastic$lambda.min)

# Predict
predictions_lasso = cv_final_lasso %>% predict(test_mat[,-6]) %>% as.vector()
predictions_ridge = cv_final_ridge %>% predict(test_mat[,-6]) %>% as.vector()
predictions_elastic = cv_final_elastic %>% predict(test_mat[,-6])%>% 
  as.vector()

# Output results
data.frame(RMSE = rmse(predictions_lasso, test_mat[,6]), 
           MAE = mae(predictions_lasso, test_mat[,6]))
data.frame(RMSE = rmse(predictions_ridge, test_mat[,6]), 
           MAE = mae(predictions_ridge, test_mat[,6]))
data.frame(RMSE = rmse(predictions_elastic, test_mat[,6]), 
           MAE = mae(predictions_elastic, test_mat[,6]))
```


## Follow-up Study:

### Discarding Age Variable


```{r}
# Build model without age
lm.no_age = lm(BPXSY1~RIAGENDR+BPXDI1+BPXML1, data=nhanes_ht_ready)
summary(lm.no_age)

# Train and predict
lm.train.no_age = lm(BPXSY1~RIAGENDR+BPXDI1+BPXML1, data=training_data)
predictions_no_age = predict(lm.train, testing_data)
data.frame( RMSE = rmse(predictions_no_age, testing_data$BPXSY1),
            MAE = mae(predictions_no_age, testing_data$BPXSY1))
```


### Effect of BP Medications


```{r, fig.height = 8, fig.width = 10}
# Separate the cohort by medication use
ht_med = nhanes_ht_ready %>% filter(BPQ100D==1)
ht_med_no = nhanes_ht_ready %>% filter(BPQ100D==0)
# Plot distribution
hist(ht_med$BPXSY1, 
     main="Distribution of Systolic BP for Group with Medication")
hist(ht_med_no$BPXSY1, 
     main="Distribution of Systolic BP for Group with No Medication")
```

```{r, fig.height = 8, fig.width = 10}
library(ggplot2)
# Plot out diastolic BP and systolic BP
ggplot(data=nhanes_ht_ready)+geom_point(aes(BPXDI1, BPXSY1, 
                                            color=as.factor(BPQ100D)))+
  labs(title="Diastolic BP vs Systolic BP")
```

```{r}
# Conduct a pearson correlation test
cor.test(nhanes_ht_ready$BPQ100D, nhanes_ht_ready$BPXSY1)
```

```{r}
# Assess the linear relationship between systolic BP and
# medication use alone
lm.bp = lm(BPXSY1~BPQ100D,data=nhanes_ht_ready)
summary(lm.bp)
```

